---
title: "hickey v cheltenham deaths"
author: "Wells"
date: "2026-02-26"
output: html_document
---

### Load libraries

```{r}
# install.packages("htmlwidgets")
# install.packages("leaflet")
#install.packages("sf")
#install.packages('DT')
library(DT)
library(tidyverse)
library(rio)
library(janitor)
library(htmlwidgets)
library(leaflet)
library(sf)
library(googlesheets4)
library(formattable)
```

#death certificates

```{r}
library(googlesheets4)

gs4_deauth()

death_certs <- read_sheet("https://docs.google.com/spreadsheets/d/1Qu7FSUzCAeU4gdZ9BpIvfxI5kBp5rAOgEiaQ2h0ff4k/edit?gid=0#gid=0") |> 
     clean_names()

chelt_deaths <- read_sheet("https://docs.google.com/spreadsheets/d/1qy17Gwxyb-W6P2tDqaI6jUymK4n7cmnFToNIkJ5GCI4/edit?usp=sharing", sheet="chelt_deaths") |> 
     clean_names() |> 
  mutate(across(
    # Select all the list columns
    c(number_of_inmates, total_deaths, total_escapes, total_hired_out, total_sent_to_hospital),
    ~ .x |> 
      as.character() |>              # Pull it out of the 'list' box into a string
      str_remove_all("\\*") |>       # Remove those asterisks (e.g., "56*" becomes "56")
      na_if("NA") |>                 # Convert the string "NA" to a real R NA
      na_if("na") |>                 # Handle lowercase "na"
      na_if("N/A") |>                # Handle "N/A"
      as.numeric()                   # Finally, turn it into a number
  )) |> 
   slice(-43)



hickey_deaths <- read_sheet("https://docs.google.com/spreadsheets/d/1FnXL2BiktaoEQDh-TAzgdTDuucJfaENKQ1wVZaItdww/edit?usp=sharing") |> 
     clean_names() |> 
    mutate(across(
    # Select all the list columns
    c(year_of_report,number_of_inmates, total_deaths, total_escapes, total_hired_out, total_sent_to_hospital),
    ~ .x |> 
      as.character() |>              # Pull it out of the 'list' box into a string
      str_remove_all("\\*") |>       # Remove those asterisks (e.g., "56*" becomes "56")
      na_if("NA") |>                 # Convert the string "NA" to a real R NA
      na_if("na") |>                 # Handle lowercase "na"
      na_if("N/A") |>                # Handle "N/A"
      as.numeric()                   # Finally, turn it into a number
  ))

```

#fact check
```{r}
dc_total <- death_certs |> 
  filter(facility =="cheltenham") |> 
  group_by(death_year) |> 
  count(, name="death_cert_total") 

dc_deaths <- sum(dc_total$death_cert_total, na.rm=TRUE) 

cat("Total death certificate deaths:", dc_deaths, "\n")

report_deaths <-  sum(chelt_deaths$total_deaths, na.rm=TRUE)

cat("Total report deaths:", report_deaths, "\n")
```


#delete this after cleaning up the code below
```{r}
chelt_death_cert_totals <- death_certs |> 
  filter(facility =="cheltenham") |> 
  group_by(death_year) |> 
  count(, name="death_cert_total") |> 
  arrange(desc(death_year)) |> 
  left_join(chelt_deaths, by=c("death_year"="year_of_report")) |> 
  select(death_year, total_deaths,  death_cert_total) |> 
  mutate(death_certs_bigger = if_else(death_cert_total > total_deaths, "Yes", "No"))

#this join keeps all death cert and all annual report entries

```


# I dont understand why you did this below - seems to duplicate the above
```{r}
chelt_death_cert <- death_certs %>% 
    select(death_year,facility) %>% 
  filter(facility =="cheltenham") %>% 
  arrange(desc(death_year))

chelt_death_cert_count <- chelt_death_cert %>% 
  group_by(death_year) %>% 
  count()
chelt_death_cert_count <-chelt_death_cert_count %>% 
rename( year_of_report =death_year)
 
  
```

```{r}
updated_chelt_death_cert <- chelt_deaths %>%
  full_join(chelt_death_cert_count, by = "year_of_report")

joined_chelt <- updated_chelt_death_cert %>% 
  select(year_of_report,total_deaths,n)

```

```{r}
chelt_deaths <- chelt_deaths %>%
  filter(!is.na(year_of_report))

updated_chelt_death_cert <- chelt_deaths %>%
  full_join(chelt_death_cert_count, by = "year_of_report")

joined_chelt <- updated_chelt_death_cert %>% 
  select(year_of_report, total_deaths, n) %>%
  rename(death_cert = n) %>%
  rename(report_deaths = total_deaths) %>% 
  mutate(larger = case_when(
    is.na(report_deaths) | is.na(death_cert) ~ "missing data",
    report_deaths > death_cert ~ "report_deaths",
    death_cert > report_deaths ~ "death_cert",
    report_deaths == death_cert ~ "equal"
  ))
```

```{r}
joined_chelt <- joined_chelt %>%
  mutate(max_deaths = case_when(
    is.na(report_deaths) & is.na(death_cert) ~ NA_real_,
    is.na(report_deaths) ~ death_cert,
    is.na(death_cert) ~ report_deaths,
    report_deaths >= death_cert ~ report_deaths,
    death_cert > report_deaths ~ death_cert
  ))

total_sum <- sum(joined_chelt$max_deaths, na.rm = TRUE)
```

```{r}
joined_chelt <- joined_chelt %>%
  mutate(discrepancy = abs(report_deaths - death_cert))
```

# Total: 258
```{r}

joined_chelt |> 
  summarize(sum(max_deaths, na.rm=TRUE))


```

Need to account for the biennial years

Table of missing years
```{r}
missing <- joined_chelt %>%
  filter(is.na(report_deaths)) %>%
  select(year_of_report)

biennial_years  <- c(1903, 1907, 1909, 1911, 1913, 1915, 1917, 1930, 1932, 1934, 1936, 1938) 

missing1 <- missing %>%
  mutate(
    # Check if the year itself is a biennial report year
    is_report_year = year_of_report %in% biennial_years,
    
    # Check if the data belongs to a biennial cycle (the year + 1)
    is_covered_by_biennial = (year_of_report %in% biennial_years) | ((year_of_report + 1) %in% biennial_years),
    
    # Label them for clarity
    status = case_when(
      year_of_report %in% biennial_years ~ "Biennial Report Year",
      (year_of_report + 1) %in% biennial_years ~ paste("Covered by", year_of_report + 1, "Report"),
      TRUE ~ "Standalone/Missing"
    )
  )

write_csv(missing1, "missing_years_analysis.csv")


missing_pia <- missing1 |> 
  filter(status == "Standalone/Missing") |> 
  arrange(year_of_report) |> 
  select(year_of_report, status) 

write_csv(missing_pia, "missing_reports_pia_request.csv")


```




#Claude

```{r}
biennial_years   <- c(1903, 1907, 1909, 1911, 1913, 1915, 1917, 1930, 1932, 1934, 1936, 1938)
biennial_prev_years <- biennial_years - 1   # years already "absorbed" into the biennial total

# --- cert counts for the absorbed previous years ---
prev_year_certs <- joined_chelt %>%
  filter(year_of_report %in% biennial_prev_years) %>%
  transmute(
    biennial_year  = year_of_report + 1,   # maps 1906 -> 1907, etc.
    cert_prev_year = death_cert
  )

# --- main calculation ---
deaths_summary <- joined_chelt %>%
  mutate(
    is_biennial  = year_of_report %in% biennial_years,
    is_prev_year = year_of_report %in% biennial_prev_years
  ) %>%
  # drop rows already covered inside a biennial report
  filter(!is_prev_year) %>%
  left_join(prev_year_certs, by = c("year_of_report" = "biennial_year")) %>%
  mutate(
    # for biennial years: combine both years' certs; for others: use as-is
    cert_total = case_when(
      is_biennial ~ coalesce(as.numeric(death_cert), 0) +
                    coalesce(as.numeric(cert_prev_year), 0),
      TRUE        ~ as.numeric(death_cert)
    ),
    # if biennial but *both* cert years are truly absent, keep NA (don't use 0)
    cert_total = if_else(
      is_biennial & is.na(death_cert) & is.na(cert_prev_year),
      NA_real_, cert_total
    ),
    # best estimate per row: take whichever source is larger
    best_estimate = case_when(
      !is.na(report_deaths) & !is.na(cert_total) ~ pmax(report_deaths, cert_total),
      !is.na(report_deaths)                       ~ report_deaths,
      !is.na(cert_total)                          ~ cert_total,
      TRUE                                         ~ NA_real_
    )
  )

total_deaths <- sum(deaths_summary$best_estimate, na.rm = TRUE)
missing_rows  <- sum(is.na(deaths_summary$best_estimate))

cat("Total deaths:", total_deaths, "\n")
cat("Rows with no data from either source:", missing_rows, "\n")

```

Claude explanation:
Looking at the problem: for biennial years (e.g. 1907), `report_deaths` covers **two** years (1906+1907), but `death_cert` in `joined_chelt` only reflects the single report year. So you need to:

1. Pull the cert counts for both constituent years of each biennial pair
2. Sum them before comparing against the biennial report total
3. Exclude the "previous year" rows from separate counting (they're already rolled into the biennial)

```r
biennial_years   <- c(1903, 1907, 1909, 1911, 1913, 1915, 1917, 1930, 1932, 1934, 1936, 1938)
biennial_prev_years <- biennial_years - 1   # years already "absorbed" into the biennial total

# --- cert counts for the absorbed previous years ---
prev_year_certs <- joined_chelt %>%
  filter(year_of_report %in% biennial_prev_years) %>%
  transmute(
    biennial_year  = year_of_report + 1,   # maps 1906 -> 1907, etc.
    cert_prev_year = death_cert
  )

# --- main calculation ---
deaths_summary <- joined_chelt %>%
  mutate(
    is_biennial  = year_of_report %in% biennial_years,
    is_prev_year = year_of_report %in% biennial_prev_years
  ) %>%
  # drop rows already covered inside a biennial report
  filter(!is_prev_year) %>%
  left_join(prev_year_certs, by = c("year_of_report" = "biennial_year")) %>%
  mutate(
    # for biennial years: combine both years' certs; for others: use as-is
    cert_total = case_when(
      is_biennial ~ coalesce(as.numeric(death_cert), 0) +
                    coalesce(as.numeric(cert_prev_year), 0),
      TRUE        ~ as.numeric(death_cert)
    ),
    # if biennial but *both* cert years are truly absent, keep NA (don't use 0)
    cert_total = if_else(
      is_biennial & is.na(death_cert) & is.na(cert_prev_year),
      NA_real_, cert_total
    ),
    # best estimate per row: take whichever source is larger
    best_estimate = case_when(
      !is.na(report_deaths) & !is.na(cert_total) ~ pmax(report_deaths, cert_total),
      !is.na(report_deaths)                       ~ report_deaths,
      !is.na(cert_total)                          ~ cert_total,
      TRUE                                         ~ NA_real_
    )
  )

total_deaths <- sum(deaths_summary$best_estimate, na.rm = TRUE)
missing_rows  <- sum(is.na(deaths_summary$best_estimate))

cat("Total deaths:", total_deaths, "\n")
cat("Rows with no data from either source:", missing_rows, "\n")
```

**Key design decisions:**

| Situation | Logic |
|---|---|
| Biennial year, both sources present | `pmax(report_deaths, cert_curr + cert_prev)` |
| Biennial year, only report | use `report_deaths` |
| Biennial year, both cert years missing | `NA` (not 0) |
| Regular year | `pmax(report_deaths, death_cert)` ignoring whichever is `NA` |
| Previous year of a biennial (e.g. 1906) | **filtered out** — already counted in the 1907 row |

The `coalesce(..., 0)` on the cert sum treats a missing cert year as 0 *only when the other cert year has data*; that's intentional — a missing cert year for a biennial where the other year is present means partial coverage, not complete absence.


# Afro Deaths
total deaths reported in the Afro and Sun, and align those with these total above.


```{r}
afro_deaths <- read_sheet("https://docs.google.com/spreadsheets/d/1375Kt-61NU1FFV4pshCEv0EV3ZN4x6dWM31_qKGHxzo/edit?usp=sharing") |> 
     clean_names() |> 
  mutate(date_clean = ymd(date_of_article),
        year = year(date_clean)) |> 
  rename(deaths = is_this_article_about_a_death_yes_or_no)

afro_deaths |> 
    filter(deaths =="Yes it's about a death") |> 
  select(year, any_notes_on_the_death_description_if_none_n_a, date_clean, deaths) |> 
  arrange(year)

```

```{r}
afro_deaths_summary <- afro_deaths %>%
  filter(deaths =="Yes it's about a death") |> 
  group_by(year) |> 
  filter(!year %in% c(1899, 1940)) |> 
  count(, name="afro_deaths") |> 
  mutate(afro_deaths = if_else(year == 1929, 1, afro_deaths))

#cut
# 1899 was about a murder that sent a kid to cheltenham
# 1929 repeated. 
# 1940 was an obit of a cheltenham manager

afro_total <- sum(afro_deaths_summary$afro_deaths, na.rm=TRUE) 

cat("Total Cheltenham deaths reported by the Afro:", afro_total, "\n")

```






https://github.com/ejalil/cheltehnam_jalil/blob/main/hickey_v_chelteham_EJ_RSW.Rmd


